---
title: "Getting Started with rOFI"
author: "Dionysios Poniros"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rOFI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r load-package}
library(rOFI)
library(dplyr)
library(ggplot2)
```

## Introduction

Order-flow imbalance (OFI) is a fundamental concept in market microstructure that measures the difference between buying and selling pressure in financial markets. The `rOFI` package makes these concepts accessible through simple, well-documented functions.

### What is Order-Flow Imbalance?

At its core, OFI answers a simple question: "Are there more buyers or sellers in the market right now?"

- **Positive OFI**: More buying than selling (bullish pressure)
- **Negative OFI**: More selling than buying (bearish pressure)  
- **Near-zero OFI**: Balanced market

## Your First OFI Analysis

Let's start with a complete example:

```{r first-example}
# Generate synthetic trading data
trades <- simulate_orders(
  n = 5000,        # 5000 trades
  lambda = 10,     # ~10 trades per minute
  imb = 0.1,       # Slight buying pressure
  seed = 123       # For reproducibility
)

# Look at the data structure
head(trades)

# Compute OFI with 5-minute windows
ofi_results <- compute_ofi(trades, window = "5 min")

# Examine the results
head(ofi_results)
```

### Understanding the Output

The `compute_ofi()` function returns several metrics:

- `window_start`: Beginning of each time window
- `B`: Total buy volume in the window
- `S`: Total sell volume in the window
- `ofi`: Simple order-flow imbalance (B - S)
- `oir`: Order imbalance ratio, normalized to [-1, 1]
- `ofi_cum`: Cumulative OFI (running total)
- `vol_total`: Total volume (B + S)
- `n_trades`: Number of trades in the window

## Visualization

The package includes built-in plotting functions:

```{r basic-plot, fig.height=6}
# Plot multiple metrics
plot_ofi(ofi_results, which = c("ofi", "oir", "ofi_cum"))
```

### Customizing Plots

```{r custom-plot}
# Focus on the imbalance ratio with custom styling
plot_ofi(ofi_results, 
         which = "oir",
         ref_line = 0,
         title = "Order Imbalance Ratio Over Time",
         subtitle = "Detecting buying and selling pressure")
```

### Distribution Analysis

Understanding the distribution of imbalance helps identify market regimes:

```{r distribution}
# Histogram of imbalance ratios
plot_ofi_dist(ofi_results, metric = "oir", bins = 25)
```

## Window Types

The package supports three types of aggregation windows:

### 1. Calendar Windows (Default)

Fixed time intervals aligned to clock time:

```{r calendar-windows}
# Different time scales
ofi_1min <- compute_ofi(trades, window = "1 min")
ofi_10min <- compute_ofi(trades, window = "10 min")
ofi_30min <- compute_ofi(trades, window = "30 min")

# Compare granularities
data.frame(
  Window = c("1 min", "10 min", "30 min"),
  Observations = c(nrow(ofi_1min), nrow(ofi_10min), nrow(ofi_30min)),
  Mean_OFI = c(mean(ofi_1min$ofi), mean(ofi_10min$ofi), mean(ofi_30min$ofi))
)
```

### 2. Rolling Windows

Overlapping windows for smoother analysis:

```{r rolling-windows}
# 60-second rolling window
ofi_rolling <- compute_ofi(trades, rolling = lubridate::dseconds(60))

# More observations due to overlap
nrow(ofi_rolling)

# Smoother patterns
plot_ofi(ofi_rolling, which = "oir", title = "Rolling 60-Second OIR")
```

### 3. Tick-Based Windows

Fixed number of trades per window (useful for irregular data):

```{r tick-windows}
# Every 100 trades
ofi_ticks <- compute_ofi(trades, n_ticks = 100)

# Consistent trade counts
summary(ofi_ticks$n_trades)

plot_ofi(ofi_ticks, which = "ofi", title = "OFI by 100-Trade Windows")
```

## Working with Real Data

The `as_ofi()` function standardizes various data formats:

```{r real-data}
# Example: Your data might look different
raw_trades <- data.frame(
  time = Sys.time() + cumsum(rexp(100, rate = 0.1)),
  direction = sample(c("BUY", "SELL"), 100, replace = TRUE, prob = c(0.55, 0.45)),
  volume = round(rlnorm(100, log(500), 0.5)),
  px = 100 + cumsum(rnorm(100, 0, 0.1))
)

# Standardize for rOFI
clean_trades <- as_ofi(
  raw_trades,
  time_col = "time",
  side_col = "direction",
  size_col = "volume",
  price_col = "px",
  tz = "America/New_York"
)

# Now ready for analysis
ofi_real <- compute_ofi(clean_trades, window = "1 min")
```

## Dollar-Weighted OFI

When prices are available, compute value-weighted metrics:

```{r dollar-ofi}
# Generate data with realistic price movement
trades_with_price <- simulate_orders(
  n = 2000,
  price0 = 100,
  drift = 0.01,    # Slight upward drift
  vol = 0.5,       # Moderate volatility
  seed = 456
)

# Compute with price weighting
ofi_dollar <- compute_ofi(trades_with_price, 
                          window = "5 min",
                          price_weighted = TRUE)

# Dollar OFI is now available
head(select(ofi_dollar, window_start, ofi, ofi_dollar))

# Compare size-weighted vs dollar-weighted
plot_ofi(ofi_dollar, which = c("ofi", "ofi_dollar"))
```

## Market Regime Detection

Use OFI patterns to identify market conditions:

```{r regime-detection}
# Simulate different market regimes
set.seed(789)

# Balanced market
balanced_trades <- simulate_orders(n = 1000, imb = 0)
# Bullish market
bullish_trades <- simulate_orders(n = 1000, imb = 0.3)
# Bearish market  
bearish_trades <- simulate_orders(n = 1000, imb = -0.3)

# Compute OFI
ofi_balanced <- compute_ofi(balanced_trades, window = "5 min")
ofi_bullish <- compute_ofi(bullish_trades, window = "5 min")
ofi_bearish <- compute_ofi(bearish_trades, window = "5 min")

# Compare statistics
regime_stats <- data.frame(
  Regime = c("Balanced", "Bullish", "Bearish"),
  Mean_OIR = c(mean(ofi_balanced$oir), 
               mean(ofi_bullish$oir),
               mean(ofi_bearish$oir)),
  SD_OIR = c(sd(ofi_balanced$oir),
             sd(ofi_bullish$oir), 
             sd(ofi_bearish$oir)),
  Skew = c(moments::skewness(ofi_balanced$oir),
           moments::skewness(ofi_bullish$oir),
           moments::skewness(ofi_bearish$oir))
)

print(regime_stats)
```

## Advanced Example: Intraday Patterns

Analyze how OFI evolves throughout a trading day:

```{r intraday, fig.height=5}
# Simulate a full trading day (9:30 AM - 4:00 PM)
trading_day <- simulate_orders(
  n = 10000,
  start = as.POSIXct("2024-01-15 09:30:00", tz = "America/New_York"),
  lambda = 25,  # ~25 trades per minute
  imb = 0.05,   # Slight overall buying pressure
  seed = 2024
)

# Compute OFI
ofi_intraday <- compute_ofi(trading_day, window = "10 min")

# Add hour of day for analysis
ofi_intraday <- ofi_intraday %>%
  mutate(hour = lubridate::hour(window_start))

# Average OFI by hour
hourly_pattern <- ofi_intraday %>%
  group_by(hour) %>%
  summarise(
    mean_oir = mean(oir),
    mean_volume = mean(vol_total),
    .groups = "drop"
  )

# Plot hourly patterns
ggplot(hourly_pattern, aes(x = hour, y = mean_oir)) +
  geom_col(fill = "#2E86AB", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Average Order Imbalance by Hour",
       x = "Hour of Day",
       y = "Mean OIR") +
  scale_x_continuous(breaks = 9:15, labels = paste0(9:15, ":00"))
```

## Performance Considerations

For large datasets, consider these optimizations:

```{r performance, eval=FALSE}
# For very large datasets (>1M trades)
large_trades <- simulate_orders(n = 1000000)

# Use larger windows to reduce output size
ofi_large <- compute_ofi(large_trades, window = "10 min")

# Or use tick-based windows for consistent computation
ofi_ticks_large <- compute_ofi(large_trades, n_ticks = 1000)

# For parallel processing (not included in base package)
library(furrr)
plan(multisession, workers = 4)

# Split data and process in parallel
# (Example code - would need custom implementation)
```

## Tips and Best Practices

1. **Choose appropriate window sizes**: 
   - High-frequency data: 1-5 minutes
   - Daily analysis: 10-30 minutes
   - Long-term trends: 1 hour or more

2. **Normalize with OIR for comparisons**: The ratio is scale-independent

3. **Check for data quality**:
```{r data-quality}
# Validate your data before analysis
check_data <- function(trades) {
  list(
    n_rows = nrow(trades),
    n_buys = sum(trades$side == "B"),
    n_sells = sum(trades$side == "S"),
    pct_buys = mean(trades$side == "B") * 100,
    size_range = range(trades$size),
    time_range = range(trades$timestamp)
  )
}

check_data(trades)
```

4. **Handle sparse data carefully**: Some windows might be empty

5. **Consider market hours**: Filter to regular trading hours if needed

## Conclusion

The `rOFI` package provides a comprehensive toolkit for order-flow analysis. Key takeaways:

- **Simple interface**: Four main functions cover most use cases
- **Flexible windows**: Calendar, rolling, or tick-based aggregation
- **Educational focus**: Clear documentation and examples
- **Production ready**: Efficient implementation with proper testing

For more examples and updates, visit the [GitHub repository](https://github.com/DennisPoniros/rOFI_package).

## Session Info

```{r session}
sessionInfo()
```
