---
title: "Advanced Visualization Dashboard"
author: "rOFI Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Visualization Dashboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The **rOFI** package provides publication-quality visualizations designed for statisticians, quantitative researchers, and market microstructure analysts. This vignette demonstrates advanced plotting capabilities that go beyond basic time series plots to include:

- **Diagnostic dashboards** with ACF/PACF and Q-Q plots
- **Time series decomposition** into trend, seasonal, and irregular components
- **Market quality dashboards** showing liquidity and efficiency metrics
- **Regime detection** for identifying market states
- **Comparative analysis** across multiple assets or periods
- **Publication-ready themes** for academic papers

All visualizations emphasize statistical rigor with proper confidence bands, hypothesis tests, and professional defaults suitable for presentations and publications.

## Setup

```{r libraries}
library(rOFI)
library(patchwork)  # Required for multi-panel layouts
library(ggplot2)
library(dplyr)
```

## Generate Sample Data

We'll create realistic order flow data with different market conditions:

```{r generate_data}
set.seed(42)

# Balanced market (neutral order flow)
trades_balanced <- simulate_orders(
  n = 2000,
  start = as.POSIXct("2024-01-15 09:30:00", tz = "America/New_York"),
  lambda = 10,    # 10 trades per minute
  imb = 0,        # No imbalance
  drift = 0,
  vol = 0.15,
  seed = 42
)

# Bullish market (buy pressure)
trades_bullish <- simulate_orders(
  n = 2000,
  start = as.POSIXct("2024-01-15 09:30:00", tz = "America/New_York"),
  lambda = 10,
  imb = 0.3,      # 30% more buys
  drift = 0.5,    # Upward price drift
  vol = 0.15,
  seed = 43
)

# High volatility market (stress conditions)
trades_volatile <- simulate_orders(
  n = 2000,
  start = as.POSIXct("2024-01-15 09:30:00", tz = "America/New_York"),
  lambda = 20,    # Higher trading intensity
  imb = 0.1,
  drift = 0,
  vol = 0.40,     # High volatility
  seed = 44
)

# Compute OFI for each scenario
ofi_balanced <- compute_ofi(trades_balanced, window = "1 min")
ofi_bullish <- compute_ofi(trades_bullish, window = "1 min")
ofi_volatile <- compute_ofi(trades_volatile, window = "1 min")
```

## 1. Diagnostic Dashboard

The `plot_ofi_diagnostics()` function creates a comprehensive 6-panel diagnostic dashboard perfect for statistical analysis:

```{r diagnostics, fig.width=10, fig.height=10}
plot_ofi_diagnostics(
  ofi_bullish,
  metric = "oir",
  max_lag = 20,
  conf_level = 0.95,
  title = "OFI Diagnostic Dashboard: Bullish Market"
)
```

### What the Panels Show

1. **Time Series with Trend**: Raw OIR values with LOESS smoothing to identify trends
2. **ACF (Autocorrelation)**: Shows persistence in order flow imbalance
3. **PACF (Partial Autocorrelation)**: Identifies direct lagged relationships
4. **Q-Q Plot**: Tests normality assumption (departures indicate fat tails)
5. **Distribution**: Histogram with density overlay and mean reference line
6. **Summary Statistics**: Key moments (mean, SD, skewness, kurtosis)

### Interpretation

- **Significant ACF at multiple lags** indicates persistent order flow patterns (momentum)
- **Negative skewness** suggests more extreme sell-side imbalances
- **Excess kurtosis** (fat tails) indicates occasional extreme imbalances
- **Q-Q plot departures** at tails show non-normality (common in financial data)

This diagnostic suite helps researchers:
- Test assumptions for econometric models
- Identify appropriate lag structures for forecasting
- Detect anomalies and outliers
- Understand distributional properties

## 2. Time Series Decomposition

Decompose OFI into trend, seasonal (intraday patterns), and irregular components:

```{r decomposition, fig.width=8, fig.height=8}
plot_ofi_decomposition(
  ofi_balanced,
  metric = "oir",
  freq = NULL  # Auto-detect frequency
)
```

### Use Cases

- **Trend component**: Identifies sustained directional pressure
- **Seasonal component**: Reveals intraday patterns (opening/closing effects)
- **Remainder**: Isolates idiosyncratic order flow shocks

This is particularly useful for:
- Filtering out predictable patterns before forecasting
- Identifying structural breaks in market behavior
- Understanding time-of-day effects in execution

## 3. Market Quality Dashboard

Comprehensive view of market conditions combining volume, imbalance, spreads, and volatility:

```{r market_quality, fig.width=10, fig.height=8}
plot_market_quality(
  trades_volatile,
  window = "5 min",
  title = "Market Quality: High Volatility Period"
)
```

### Key Metrics

1. **Trading Volume**: Liquidity and activity level
2. **Order Imbalance Ratio**: Buy/sell pressure
3. **Spread Proxy**: Transaction cost estimate
4. **Price Volatility**: Risk and uncertainty

### Interpretation

- **Simultaneous spikes** in volume and volatility indicate news arrival
- **High OIR with expanding spreads** suggests informed trading
- **Low volume with high spreads** indicates illiquidity
- **High volatility with balanced OIR** suggests noise traders

Market makers and execution desks use these dashboards for:
- Real-time risk management
- Optimal execution timing
- Spread adjustment decisions
- Regulatory surveillance

## 4. Regime Detection

Identify distinct market states using statistical clustering:

```{r regime_quantile, fig.width=10, fig.height=6}
# Quantile-based regimes (simple thresholds)
plot_regime_detection(
  ofi_bullish,
  method = "quantile",
  n_regimes = 3
)
```

```{r regime_kmeans, fig.width=10, fig.height=6}
# K-means clustering (multivariate)
plot_regime_detection(
  ofi_bullish,
  method = "kmeans",
  n_regimes = 3,
  vars = c("oir", "vol_total")
)
```

### Methods

- **Quantile**: Simple threshold-based regimes (low/medium/high imbalance)
- **K-means**: Multivariate clustering on OFI and volume
- **HMM** (planned): Hidden Markov Models for state transitions

### Applications

- **Regime-dependent trading strategies**: Different algorithms for different states
- **Risk management**: Increase caution in extreme regimes
- **Model selection**: Use regime-specific forecasting models
- **Stress testing**: Analyze performance across regimes

Academic research uses regime detection for:
- Testing market microstructure theories in different states
- Analyzing high-frequency trader behavior by regime
- Studying market maker inventory management

## 5. Comparative Analysis

Compare OFI patterns across multiple assets, time periods, or market conditions:

```{r comparative_facet, fig.width=10, fig.height=10}
# Faceted comparison (separate panels)
plot_comparative_analysis(
  ofi_list = list(
    "Balanced Market" = ofi_balanced,
    "Bullish Market" = ofi_bullish,
    "High Volatility" = ofi_volatile
  ),
  metric = "oir",
  layout = "facet",
  normalize = FALSE
)
```

### With Standardization

For comparing markets with different scales:

```{r comparative_normalized, fig.width=10, fig.height=10}
plot_comparative_analysis(
  ofi_list = list(
    "Balanced" = ofi_balanced,
    "Bullish" = ofi_bullish,
    "Volatile" = ofi_volatile
  ),
  metric = "oir",
  layout = "facet",
  normalize = TRUE  # Z-score standardization
)
```

### Overlay Comparison

```{r comparative_overlay, fig.width=10, fig.height=6}
# Overlay on single plot (best for normalized data)
plot_comparative_analysis(
  ofi_list = list(
    "Balanced" = ofi_balanced,
    "Bullish" = ofi_bullish
  ),
  metric = "oir",
  layout = "overlay",
  normalize = TRUE
)
```

### Use Cases

**Cross-sectional studies**:
- Compare OFI across stocks (tech vs. finance vs. energy)
- Analyze ETF vs. constituent patterns
- Study futures vs. spot markets

**Temporal comparisons**:
- Pre-market vs. regular hours vs. after-hours
- Compare same stock across different dates
- Analyze before/after major news events

**Market condition studies**:
- Bull markets vs. bear markets
- High vs. low volatility regimes
- Liquid vs. illiquid periods

## 6. Publication-Ready Theme

The `theme_publication()` creates clean, professional plots suitable for academic papers:

```{r publication_theme, fig.width=8, fig.height=5}
# Apply to any plot
plot_ofi(ofi_bullish, which = "oir") +
  theme_publication() +
  labs(
    title = "Order Imbalance Ratio: Bullish Market Conditions",
    subtitle = "1-minute aggregation windows"
  )
```

### Theme Features

- **Minimalist design**: Removes visual clutter
- **Bold titles and axis labels**: Ensures readability
- **Clean grid lines**: Subtle major grid only
- **Professional fonts**: Compatible with journal requirements
- **Optimal spacing**: Publication-ready margins

### Customization

```{r publication_custom, fig.width=8, fig.height=5}
plot_ofi(ofi_balanced, which = c("ofi", "ofi_cum")) +
  theme_publication(base_size = 12) +
  labs(title = "Order Flow Imbalance Patterns") +
  scale_color_viridis_d()  # Colorblind-friendly palette
```

## 7. Creating Custom Dashboards

Combine multiple visualizations using `patchwork`:

```{r custom_dashboard, fig.width=12, fig.height=10}
# Create individual plots
p1 <- plot_ofi(ofi_bullish, which = "oir") +
  theme_publication() +
  labs(title = "A. Order Imbalance Ratio")

p2 <- plot_ofi_dist(ofi_bullish, metric = "oir", plot_type = "density") +
  theme_publication() +
  labs(title = "B. OIR Distribution")

p3 <- plot_regime_detection(ofi_bullish, method = "quantile", n_regimes = 3) +
  theme_publication() +
  labs(title = "C. Market Regimes")

# Combine into custom layout
(p1 | p2) / p3 +
  plot_annotation(
    title = "Comprehensive OFI Analysis Dashboard",
    subtitle = "Bullish Market Conditions - January 15, 2024",
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )
```

## 8. Complete Analysis Workflow

Here's a complete analysis workflow for a research paper:

```{r workflow, fig.width=12, fig.height=10}
# 1. Load or generate data
trades <- simulate_orders(n = 3000, imb = 0.15, seed = 123)
ofi <- compute_ofi(trades, window = "1 min")

# 2. Initial exploratory plots
p_timeseries <- plot_ofi(ofi, which = c("ofi", "oir", "ofi_cum")) +
  theme_publication()

# 3. Statistical diagnostics (use separately)
# diagnostics <- plot_ofi_diagnostics(ofi, metric = "oir", max_lag = 20)

# 4. Test for statistical significance
acf_test <- test_ofi_autocorrelation(ofi, max_lag = 20)
print(paste("Box-Ljung test p-value:", round(acf_test$p_value, 4)))

# 5. Regime analysis
p_regime <- plot_regime_detection(ofi, method = "kmeans", n_regimes = 3)

# 6. Distribution analysis
p_dist <- plot_ofi_dist(ofi, metric = "oir", plot_type = "histogram") +
  theme_publication()

# Show subset of workflow
p_timeseries
```

## Best Practices for Research

### For Statistical Analysis

1. **Always check diagnostics first**: Use `plot_ofi_diagnostics()` before modeling
2. **Test assumptions**: Q-Q plots for normality, ACF for independence
3. **Report confidence bands**: All plots include proper statistical bounds
4. **Document data quality**: Use validation functions before visualization

### For Publications

1. **Use `theme_publication()`**: Ensures consistent professional appearance
2. **Include sample sizes**: Add N to plot annotations
3. **Report test statistics**: Combine plots with statistical test results
4. **Use colorblind-safe palettes**: Default colors are accessibility-friendly
5. **Export high-resolution**: Use `ggsave()` with at least 300 DPI

```{r export_example, eval=FALSE}
# Export for publication
p <- plot_ofi_diagnostics(ofi, metric = "oir")

ggsave(
  filename = "ofi_diagnostics.pdf",
  plot = p,
  width = 10,
  height = 10,
  dpi = 300
)
```

### For Trading/Risk Management

1. **Monitor market quality**: Use `plot_market_quality()` for real-time dashboards
2. **Detect regime changes**: Update execution strategies based on regimes
3. **Compare across assets**: Use `plot_comparative_analysis()` for pairs trading
4. **Track historical patterns**: Decompose to separate trend from noise

## Advanced Customization

All plots return `ggplot2` objects and can be further customized:

```{r advanced_custom, fig.width=8, fig.height=5}
plot_ofi(ofi_bullish, which = "oir") +
  # Add custom reference lines
  geom_hline(yintercept = c(-0.3, 0.3),
             linetype = "dashed",
             color = "red",
             alpha = 0.6) +
  # Annotate important events
  annotate("rect",
           xmin = min(ofi_bullish$window_start) + 30*60,
           xmax = min(ofi_bullish$window_start) + 60*60,
           ymin = -Inf, ymax = Inf,
           alpha = 0.2, fill = "yellow") +
  annotate("text",
           x = min(ofi_bullish$window_start) + 45*60,
           y = 0.5,
           label = "News Event",
           angle = 90,
           size = 3) +
  # Apply publication theme
  theme_publication() +
  # Custom colors
  scale_color_manual(values = c("#1f77b4")) +
  # Labels
  labs(
    title = "Order Flow Analysis with Event Annotation",
    subtitle = "Highlighted period: 30-60 minutes after open"
  )
```

## Summary

The **rOFI** visualization dashboard provides:

✅ **Statistical rigor**: Proper confidence bands, diagnostic tests, and decompositions
✅ **Publication quality**: Professional themes and high-resolution export
✅ **Comprehensive analysis**: From diagnostics to regime detection to comparisons
✅ **Flexibility**: Full ggplot2 compatibility for customization
✅ **Practical utility**: Tools for both research and trading applications

### Key Functions

| Function | Purpose |
|----------|---------|
| `plot_ofi_diagnostics()` | 6-panel statistical dashboard (ACF, PACF, Q-Q, distribution) |
| `plot_ofi_decomposition()` | Trend/seasonal/irregular decomposition |
| `plot_market_quality()` | Volume, spreads, volatility, imbalance dashboard |
| `plot_regime_detection()` | Identify market states with clustering |
| `plot_comparative_analysis()` | Compare multiple assets/periods |
| `theme_publication()` | Professional theme for papers |

### Next Steps

- **Explore microstructure models**: See `vignette("getting-started")` for Kyle's lambda, VPIN, and spread analysis
- **Advanced OFI calculations**: Learn about EWMA, momentum, and acceleration in the introductory vignette
- **Data import**: Check `vignette("data-preparation")` for loading real market data
- **Statistical testing**: Use `test_ofi_autocorrelation()` and `ofi_lead_lag_analysis()` for rigorous hypothesis testing

## References

### Market Microstructure
- Cont, R., Kukanov, A., & Stoikov, S. (2014). "The Price Impact of Order Book Events." *Journal of Financial Econometrics*, 12(1), 47-88.
- Cartea, Á., Jaimungal, S., & Penalva, J. (2015). *Algorithmic and High-Frequency Trading*. Cambridge University Press.

### Time Series Decomposition
- Cleveland, R. B., Cleveland, W. S., McRae, J. E., & Terpenning, I. (1990). "STL: A Seasonal-Trend Decomposition Procedure Based on Loess." *Journal of Official Statistics*, 6(1), 3-73.

### Visualization Best Practices
- Tufte, E. R. (2001). *The Visual Display of Quantitative Information*. Graphics Press.
- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. O'Reilly Media.

---

**Package**: rOFI
**License**: MIT
**GitHub**: https://github.com/DennisPoniros/rOFI_package
