---
title: "Data Preparation Guide"
author: "rOFI Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Don't run examples that need external files
)
```

```{r setup}
library(rOFI)
```

## Introduction

This guide shows you how to prepare real-world trade data for analysis with `rOFI`. We'll cover:

1. Understanding data requirements
2. Loading data from various sources
3. Handling common data issues
4. Validating your data
5. Format-specific guides

## Data Requirements

For OFI analysis, you need **trade-level data** with these minimum columns:

| Column | Type | Description | Required |
|--------|------|-------------|----------|
| **Timestamp** | POSIXct/numeric | When the trade occurred | ✓ Yes |
| **Side** | character/numeric | Buy or Sell | ✓ Yes |
| **Size** | numeric | Trade size (shares/contracts) | ✓ Yes |
| **Price** | numeric | Execution price | Optional |

### Common Side Encodings

`rOFI` recognizes various ways to indicate buy/sell:

```{r side-encodings, eval=TRUE}
# These all work:
# - "B" / "S"
# - "BUY" / "SELL"
# - "buy" / "sell"
# - 1 / -1 (numeric)
# - "BID" / "ASK"
```

## Quick Data Preview

Before loading large files, preview them first:

```{r preview}
# Preview any CSV file
preview_trade_file("my_trades.csv", n_rows = 20)
```

This shows:
- Column names and types
- First few rows
- Detected timestamp and side columns
- Suggested read command

## Loading Common Data Formats

### 1. Simple CSV with Headers

**Example file (`trades.csv`):**
```
timestamp,side,size,price
2024-01-15 09:30:01,B,100,150.25
2024-01-15 09:30:02,S,200,150.24
2024-01-15 09:30:03,B,150,150.26
```

**Load it:**
```{r simple-csv}
trades <- read_trade_csv(
  "trades.csv",
  time_col = "timestamp",
  side_col = "side",
  size_col = "size",
  price_col = "price"
)
```

### 2. CSV Without Headers

**Example file:**
```
1642248601,1,100,150.25
1642248602,-1,200,150.24
1642248603,1,150,150.26
```

**Load it:**
```{r no-header}
trades <- read_trade_csv(
  "trades_noheader.csv",
  time_col = 1,          # First column
  side_col = 2,          # Second column
  size_col = 3,          # Third column
  price_col = 4,         # Fourth column
  time_format = "unix",  # Unix timestamp
  col_names = FALSE      # No header row
)
```

### 3. Intraday Data (Time-Only)

**Example file:**
```
time,direction,volume,px
09:30:01,BUY,100,150.25
09:30:02,SELL,200,150.24
09:30:03,BUY,150,150.26
```

**Load it:**
```{r intraday}
trades <- read_trade_csv(
  "intraday.csv",
  time_col = "time",
  side_col = "direction",
  size_col = "volume",
  price_col = "px",
  time_format = "time_only",
  date = "2024-01-15",  # Supply the date
  tz = "America/New_York"
)
```

### 4. Custom Side Labels

**If your data uses non-standard labels:**
```{r custom-sides}
trades <- read_trade_csv(
  "trades.csv",
  time_col = "time",
  side_col = "direction",
  size_col = "quantity",
  price_col = "price",
  side_mapping = c(
    "BID" = "B",
    "ASK" = "S"
  )
)
```

## Academic Data Formats

### LOBSTER (Limit Order Book Data)

[LOBSTER](https://lobsterdata.com/) is a popular academic dataset. It provides high-frequency limit order book data.

**File structure:**
- Message file: `TICKER_DATE_START_END_message_N.csv`
- Orderbook file: `TICKER_DATE_START_END_orderbook_N.csv`

**Load LOBSTER trades:**
```{r lobster}
# Read AAPL trades from January 15, 2024
trades <- read_lobster_trades(
  file_path = "AAPL_2024-01-15_34200000_57600000_message_10.csv",
  date = "2024-01-15",
  tz = "America/New_York"
)

# Automatically filters to trade executions (event types 4 and 5)
# Converts LOBSTER format to rOFI standard format
```

**What you get:**
```{r lobster-output, eval=TRUE, echo=FALSE}
cat("✓ Successfully loaded LOBSTER data
  - timestamp: POSIXct in specified timezone
  - side: 'B' or 'S'
  - size: shares traded
  - price: execution price
")
```

## Data Validation

Always validate your data after loading:

```{r validate}
# Comprehensive validation
validation <- validate_trade_data(trades)
print(validation)
```

**The validator checks for:**
- Missing timestamps or sizes
- Duplicate timestamps
- Non-chronological order
- Invalid side values
- Negative or zero sizes/prices
- Extreme outliers
- Timezone consistency

### Interpreting Validation Results

```{r validation-interpret, eval=TRUE}
# Example output:
cat("
Trade Data Validation Results
==============================

Status: FAILED

Issues detected:
  - 5 missing timestamps (0.5%)
  - 12 duplicate timestamps (1.2%)
  - Timestamps are not in chronological order
  - 3 negative sizes

Summary Statistics:
  Rows: 1000
  Time range: 2024-01-15 09:30:00 to 2024-01-15 16:00:00
  Duration: 6.5 hours
  Size: mean = 250.3, median = 180.0
  Buy percentage: 52.3%
")
```

## Data Cleaning

If validation finds issues, clean your data:

```{r clean}
# Automated cleaning
trades_clean <- clean_trade_data(
  trades,
  remove_outliers = TRUE,
  remove_duplicates = TRUE,
  outlier_method = "iqr"
)

# Check cleaning report
attr(trades_clean, "cleaning_report")
```

**Example cleaning report:**
```{r cleaning-report, eval=TRUE}
cat("
Cleaning Report:
  Original rows: 1000
  Removed missing: 5
  Removed invalid size: 3
  Removed duplicates: 12
  Removed outliers: 8
  Final rows: 972
  Retention rate: 97.2%
")
```

## Handling Time Zones

Time zones are critical for multi-market analysis:

```{r timezones}
# Specify timezone when loading
trades_nyse <- read_trade_csv(
  "nyse_trades.csv",
  tz = "America/New_York",
  time_col = 1, side_col = 2, size_col = 3, price_col = 4
)

trades_london <- read_trade_csv(
  "lse_trades.csv",
  tz = "Europe/London",
  time_col = 1, side_col = 2, size_col = 3, price_col = 4
)

# Convert if needed
library(lubridate)
trades_nyse_utc <- trades_nyse
trades_nyse_utc$timestamp <- with_tz(trades_nyse$timestamp, "UTC")
```

## Common Data Issues & Solutions

### Issue 1: Mixed Timezones

```{r mixed-tz}
# Check current timezone
attr(trades$timestamp, "tzone")

# Standardize to UTC
library(lubridate)
trades$timestamp <- with_tz(trades$timestamp, "UTC")
```

### Issue 2: Inconsistent Side Labels

```{r inconsistent-sides}
# Check unique values
unique(trades$side)
# [1] "B" "S" "BUY" "SELL"  # Mixed!

# Standardize with as_ofi()
trades <- as_ofi(
  trades,
  side_map = c(
    "B" = "B", "BUY" = "B",
    "S" = "S", "SELL" = "S"
  )
)
```

### Issue 3: Large Files

```{r large-files}
# Read in chunks for very large files
library(readr)

# Method 1: Use data.table for speed
library(data.table)
trades <- fread("huge_file.csv")
trades <- as.data.frame(trades)  # Convert to data frame

# Method 2: Use arrow for out-of-memory processing
library(arrow)
trades <- read_csv_arrow("huge_file.csv")
```

### Issue 4: Missing Prices

```{r missing-prices}
# OFI doesn't require prices for basic analysis
trades_no_price <- trades[, c("timestamp", "side", "size")]

# This works fine:
ofi <- compute_ofi(trades_no_price, window = "1 min")

# But you can't use price_weighted = TRUE
```

## Data Quality Checklist

Before running OFI analysis, verify:

```{r checklist, eval=TRUE}
cat("
✓ Data Quality Checklist:

1. [ ] All required columns present (timestamp, side, size)
2. [ ] Timestamps are POSIXct with correct timezone
3. [ ] Side values are 'B' or 'S' (or will be mapped)
4. [ ] No missing values in critical columns
5. [ ] Timestamps are chronologically ordered
6. [ ] No negative or zero sizes
7. [ ] Data covers expected time period
8. [ ] Outliers identified and handled
9. [ ] Duplicate timestamps resolved
10. [ ] Sample data looks reasonable
")
```

## Complete Example Workflow

```{r complete-example}
# 1. Preview the file
preview_trade_file("my_data.csv")

# 2. Load with appropriate settings
trades_raw <- read_trade_csv(
  "my_data.csv",
  time_col = "timestamp",
  side_col = "side",
  size_col = "volume",
  price_col = "price",
  tz = "America/New_York"
)

# 3. Validate
validation <- validate_trade_data(trades_raw)
print(validation)

# 4. Clean if needed
if (!validation$valid) {
  trades <- clean_trade_data(trades_raw)
} else {
  trades <- trades_raw
}

# 5. Verify cleaned data
validate_trade_data(trades)

# 6. Ready for analysis!
ofi <- compute_ofi(trades, window = "1 min")
plot_ofi(ofi)
```

## Format-Specific Guides

### Bloomberg Terminal Export

Bloomberg data typically exports as:
```
Time,Side,Size,Price
09:30:01,BID,100,150.25
09:30:02,ASK,200,150.24
```

```{r bloomberg}
trades <- read_trade_csv(
  "bloomberg_export.csv",
  time_col = "Time",
  side_col = "Side",
  size_col = "Size",
  price_col = "Price",
  time_format = "time_only",
  date = "2024-01-15",
  tz = "America/New_York",
  side_mapping = c("BID" = "B", "ASK" = "S")
)
```

### Interactive Brokers (IB)

IB TWS exports often use:
```
Date,Time,Type,Size,Price
2024-01-15,09:30:01,BOT,100,150.25
2024-01-15,09:30:02,SLD,200,150.24
```

```{r ib}
# Read and combine date+time
ib_data <- read.csv("ib_trades.csv")

# Create combined timestamp
library(lubridate)
ib_data$timestamp <- ymd_hms(
  paste(ib_data$Date, ib_data$Time),
  tz = "America/New_York"
)

# Standardize
trades <- as_ofi(
  ib_data,
  time_col = "timestamp",
  side_col = "Type",
  size_col = "Size",
  price_col = "Price",
  side_map = c("BOT" = "B", "SLD" = "S")
)
```

## Troubleshooting

### Error: "Could not parse datetime"

```{r parse-error}
# Try specifying custom format
# For "15/01/2024 09:30:01" format:
trades <- read_trade_csv(
  file,
  time_col = "timestamp",
  time_format = "datetime",  # Will try multiple formats
  # ...
)

# If that fails, parse manually:
data <- read.csv(file)
library(lubridate)
data$timestamp <- dmy_hms(data$time_column, tz = "America/New_York")
trades <- as_ofi(data)
```

### Error: "Unknown side value"

```{r side-error}
# Check what values are in your data
unique(your_data$side_column)

# Provide explicit mapping
trades <- read_trade_csv(
  file,
  side_mapping = c(
    "YOUR_BUY_VALUE" = "B",
    "YOUR_SELL_VALUE" = "S"
  )
)
```

### Warning: "X rows with missing timestamps"

```{r missing-error}
# Option 1: Remove missing rows (automatic with clean_trade_data)
trades <- clean_trade_data(raw_data)

# Option 2: Investigate why they're missing
missing_idx <- which(is.na(raw_data$timestamp))
problem_rows <- raw_data[missing_idx, ]
View(problem_rows)
```

## Best Practices

1. **Always preview first**: Use `preview_trade_file()` before loading large files

2. **Validate early**: Run `validate_trade_data()` immediately after loading

3. **Keep raw data**: Save original data separate from cleaned data
   ```{r best-practice}
   trades_raw <- read_trade_csv("data.csv")
   trades_clean <- clean_trade_data(trades_raw)
   # Keep both!
   ```

4. **Document your cleaning**: Record what changes were made
   ```{r document}
   cleaning_report <- attr(trades_clean, "cleaning_report")
   saveRDS(cleaning_report, "cleaning_log.rds")
   ```

5. **Check sample periods**: Verify data makes sense
   ```{r sample-check}
   # Check first hour
   first_hour <- trades[1:100, ]
   summary(first_hour)

   # Visualize sample
   sample_ofi <- compute_ofi(first_hour, window = "1 min")
   plot_ofi(sample_ofi)
   ```

## Getting Help

If you're stuck:

```{r help}
# Function help
?read_trade_csv
?read_lobster_trades
?validate_trade_data

# Report data loading issues
# https://github.com/DennisPoniros/rOFI_package/issues
```

## Summary

**Key Steps:**
1. **Preview** your data file
2. **Load** with appropriate format settings
3. **Validate** to catch issues
4. **Clean** problematic data
5. **Verify** cleaned data looks good
6. **Analyze** with confidence!

Happy data wrangling! 📁➡️📊
