---
title: "Introduction to Order-Flow Imbalance (OFI)"
author: "rOFI Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Order-Flow Imbalance (OFI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is Order-Flow Imbalance?

If you're new to market microstructure or quantitative finance, welcome! This guide will introduce you to **Order-Flow Imbalance (OFI)** - a powerful concept for understanding how buy and sell pressure affects prices.

### The Basic Idea

Imagine a simple market where people are buying and selling a stock:

- When there are **more buyers** than sellers → price tends to go **up**
- When there are **more sellers** than buyers → price tends to go **down**

**Order-Flow Imbalance** measures this buying vs. selling pressure by looking at actual trades.

### A Simple Analogy

Think of OFI like a tug-of-war:

```
Sellers ←--------●--------→ Buyers
        (negative OFI)  (positive OFI)
```

- **Positive OFI**: More buying pressure (rope pulled toward buyers)
- **Negative OFI**: More selling pressure (rope pulled toward sellers)
- **Zero OFI**: Balanced (no one winning)

## Installing and Loading rOFI

```{r setup, eval=FALSE}
# Install from GitHub (if not already installed)
# devtools::install_github("DennisPoniros/rOFI_package")

# Load the package
library(rOFI)
```

```{r actual-setup, include=FALSE}
library(rOFI)
```

## Your First OFI Analysis

Let's start with a complete example from data generation to visualization.

### Step 1: Generate Some Trade Data

Since you might not have real market data yet, `rOFI` can create realistic synthetic data:

```{r generate-data}
# Generate 1000 trades with slight buying pressure
trades <- simulate_orders(
  n = 1000,           # Number of trades
  imb = 0.1,         # 10% more buys than sells
  seed = 123          # For reproducibility
)

# Look at the first few trades
head(trades)
```

Each trade has:
- `timestamp`: When the trade happened
- `side`: "B" for buy, "S" for sell
- `size`: How many shares traded
- `price`: Execution price

### Step 2: Calculate OFI

Now let's measure the order flow imbalance over 1-minute windows:

```{r compute-ofi}
# Compute OFI in 1-minute windows
ofi_1min <- compute_ofi(trades, window = "1 min")

# Look at the results
head(ofi_1min)
```

The output contains:
- `window_start`: Start of each time window
- `ofi`: Buy volume - Sell volume (the main metric!)
- `oir`: Order imbalance ratio (normalized to -1 to 1)
- `B`: Total buy volume
- `S`: Total sell volume
- `vol_total`: Total volume
- `n_trades`: Number of trades

### Step 3: Visualize the Results

```{r plot-ofi, fig.height=6}
# Plot the OFI over time
plot_ofi(ofi_1min, which = "ofi")
```

**What am I looking at?**

- **Positive values**: More buying than selling (bullish pressure)
- **Negative values**: More selling than buying (bearish pressure)
- **Trend**: Is buying/selling pressure increasing or decreasing?

## Understanding OFI Metrics

`rOFI` calculates several related metrics. Here's what they mean:

### 1. OFI (Order-Flow Imbalance)

**Formula**: `OFI = Buy Volume - Sell Volume`

**Example**: If buyers trade 10,000 shares and sellers trade 7,000 shares:
```
OFI = 10,000 - 7,000 = 3,000 (positive → buying pressure)
```

### 2. OIR (Order Imbalance Ratio)

**Formula**: `OIR = (Buy - Sell) / (Buy + Sell)`

**Range**: -1 to +1

**Example**: With 10,000 buy and 7,000 sell:
```
OIR = (10,000 - 7,000) / (10,000 + 7,000) = 3,000 / 17,000 ≈ 0.176
```

**Why use OIR?**
- Normalized (always between -1 and 1)
- Easier to compare across different stocks or time periods
- Less affected by overall volume changes

### 3. Cumulative OFI

**Formula**: `Sum of all OFI values up to current time`

**Purpose**: Shows the overall trend

```{r plot-cumulative, fig.height=6}
plot_ofi(ofi_1min, which = "ofi_cum")
```

**Interpretation**:
- **Rising**: Persistent buying pressure
- **Falling**: Persistent selling pressure
- **Flat**: Balanced market

## Different Time Windows

You can analyze OFI at different time scales:

```{r different-windows}
# Fast (30 seconds)
ofi_30sec <- compute_ofi(trades, window = "30 sec")

# Medium (1 minute) - we already did this
# ofi_1min <- compute_ofi(trades, window = "1 min")

# Slow (5 minutes)
ofi_5min <- compute_ofi(trades, window = "5 min")

# Compare number of windows
cat("30-second windows:", nrow(ofi_30sec), "\n")
cat("1-minute windows:", nrow(ofi_1min), "\n")
cat("5-minute windows:", nrow(ofi_5min), "\n")
```

**Trade-off**:
- **Shorter windows** → More responsive, but noisier
- **Longer windows** → Smoother, but slower to react

## Common Use Cases

### 1. Detecting Market Pressure

```{r pressure-example}
# Find windows with strong buying pressure
strong_buying <- ofi_1min[ofi_1min$oir > 0.3, ]
cat("Found", nrow(strong_buying), "periods with strong buying (OIR > 0.3)\n")

# Find windows with strong selling pressure
strong_selling <- ofi_1min[ofi_1min$oir < -0.3, ]
cat("Found", nrow(strong_selling), "periods with strong selling (OIR < -0.3)\n")
```

### 2. Trend Analysis

```{r trend-analysis}
# Is buying pressure increasing over time?
recent_avg <- mean(tail(ofi_1min$ofi, 10))
early_avg <- mean(head(ofi_1min$ofi, 10))

cat("Early OFI average:", round(early_avg, 2), "\n")
cat("Recent OFI average:", round(recent_avg, 2), "\n")

if (recent_avg > early_avg) {
  cat("→ Buying pressure is INCREASING\n")
} else {
  cat("→ Selling pressure is INCREASING\n")
}
```

### 3. Distribution Analysis

```{r distribution, fig.height=5}
# How is OFI distributed?
plot_ofi_dist(ofi_1min, metric = "oir", plot_type = "histogram")
```

**What to look for**:
- **Centered at zero**: Balanced market
- **Skewed positive**: More buying pressure overall
- **Skewed negative**: More selling pressure overall
- **Wide spread**: High variability in order flow

## Working with Your Own Data

When you have real market data, use `as_ofi()` to standardize it:

```{r own-data-example, eval=FALSE}
# Example with your own data
my_trades <- data.frame(
  timestamp = Sys.time() + 1:100,
  side = sample(c("B", "S"), 100, replace = TRUE),
  size = runif(100, 100, 1000),
  price = 100 + cumsum(rnorm(100, 0, 0.1))
)

# Standardize for rOFI
clean_data <- as_ofi(my_trades)

# Now compute OFI
ofi <- compute_ofi(clean_data, window = "1 min")
```

## Next Steps

Now that you understand the basics:

1. **Read the "Getting Started" vignette** for more advanced features
2. **Read the "Data Preparation" vignette** for loading real data
3. **Experiment** with different window sizes and parameters
4. **Explore** the statistical testing functions (autocorrelation, lead-lag analysis)

## Quick Reference

### Basic Workflow

```{r quick-ref, eval=FALSE}
# 1. Load/generate data
trades <- simulate_orders(n = 1000, seed = 123)

# 2. Compute OFI
ofi <- compute_ofi(trades, window = "1 min")

# 3. Visualize
plot_ofi(ofi)

# 4. Analyze
summary(ofi$oir)
```

### Common Parameters

```{r params-ref, eval=FALSE}
# Calendar windows
compute_ofi(trades, window = "1 min")   # 1 minute
compute_ofi(trades, window = "5 min")   # 5 minutes
compute_ofi(trades, window = "1 hour")  # 1 hour

# Tick-based (every N trades)
compute_ofi(trades, n_ticks = 100)      # Every 100 trades

# With price weighting
compute_ofi(trades, window = "1 min", price_weighted = TRUE)
```

## Getting Help

If you encounter issues:

```{r help, eval=FALSE}
# Function documentation
?compute_ofi
?plot_ofi

# Package overview
?rOFI

# Report bugs
# https://github.com/DennisPoniros/rOFI_package/issues
```

## Summary

**Key Takeaways:**

1. **OFI measures buying vs. selling pressure**
2. **Positive OFI = more buying; Negative OFI = more selling**
3. **Use different time windows** for different trading styles
4. **OIR normalizes OFI** to a -1 to +1 scale
5. **Cumulative OFI shows overall trend**

Happy analyzing! 📊
