---
title: "Testing All rOFI Functions with Synthetic Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing All rOFI Functions with Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette shows you how to test and learn **every** rOFI function using synthetic data. Perfect for:

- Learning how each module works
- Testing your code before using real data
- Teaching market microstructure concepts
- Validating your analysis pipeline

## Generate Comprehensive Test Data

The `generate_test_data()` function creates all data formats needed for rOFI:

```{r setup}
library(rOFI)

# Generate all test data formats
test_data <- generate_test_data(
  n_trades = 1000,
  n_messages = 2000,
  n_snapshots = 100,
  n_levels = 10,
  seed = 42
)

# See what you got
print(test_data)
```

## Module 1: Basic OFI Analysis

```{r module1}
# Use trade data
trades <- test_data$trades

# Compute OFI
ofi <- compute_ofi(trades, window = "1 min")
head(ofi)

# Visualize
plot_ofi(ofi, which = c("ofi", "oir"))
```

## Module 2: Advanced Visualization

```{r module2}
# Diagnostic dashboard
plot_ofi_diagnostics(ofi, metric = "oir", max_lag = 20)

# STL decomposition
plot_ofi_decomposition(ofi, metric = "oir")

# Regime detection
plot_regime_detection(ofi, method = "quantile", n_regimes = 3)
```

## Module 3: Price Impact Models

```{r module3}
# Square-root law
impact <- sqrt_impact(Q = 10000, V = 100000, sigma = 0.02)
cat("Price impact:", impact, "bps\n")

# Almgren-Chriss optimal execution
trajectory <- almgren_chriss_trajectory(
  Q = 100000,
  T_horizon = 60,
  lambda = 1e-6
)
print(trajectory)

# Calibrate from execution data
executions <- test_data$executions
calibration <- calibrate_sqrt_law(executions)
print(calibration)
```

## Module 4: Regulatory Surveillance

```{r module4}
# Use message data
messages <- test_data$messages

# Detect spoofing
spoof_result <- detect_spoofing(
  messages,
  cancel_threshold = 0.70,
  size_threshold = 0.90
)
print(spoof_result)

# Detect layering
layer_result <- detect_layering(
  messages,
  n_levels = 5,
  coordination_threshold = 0.70
)
print(layer_result)

# Multi-algorithm surveillance
alerts <- surveillance_alert_system(
  messages,
  sensitivity = "medium"
)
print(alerts)
```

## Module 5: Multi-Level Order Book

```{r module5}
# Use orderbook snapshots
orderbook <- test_data$orderbook

# Multi-level OFI
ml_ofi <- compute_multilevel_ofi(
  orderbook,
  n_levels = 5,
  weighting = "volume"
)
print(ml_ofi)

# Order book slope (depth decay)
slope <- orderbook_slope(orderbook, n_levels = 10)
print(slope)

# Microprice
mp <- microprice(orderbook, n_levels = 1)
head(mp)
```

## Module 6: Cross-Market Analysis

```{r module6}
# Use multi-venue data
multi_venue <- test_data$multi_venue

# Cross-asset OFI
cross_ofi <- compute_cross_asset_ofi(
  multi_venue,
  window = "1 min"
)
print(cross_ofi)

# Lead-lag analysis (which venue leads?)
leadlag <- lead_lag_analysis(
  cross_ofi$ofi_data$NYSE,
  cross_ofi$ofi_data$NASDAQ,
  max_lag = 10
)
print(leadlag)
```

## Module 7: Machine Learning Features

```{r module7}
# Engineer features from trade data
features <- engineer_ofi_features(
  trades,
  orderbook = orderbook,
  lookback = 5,
  n_levels = 3
)

cat("Generated", ncol(features), "features\n")
head(features[, 1:10])  # Show first 10 features

# Create event bars (better than time bars for ML)
tick_bars <- create_event_bars(
  trades,
  bar_type = "tick",
  bar_size = 100
)
cat("Created", nrow(tick_bars), "tick bars\n")

# Stationarize for modeling
stationary <- stationarize_ofi(
  ofi,
  method = "standardize"
)
cat("Mean:", mean(stationary$oir, na.rm = TRUE), "(should be ~0)\n")
cat("SD:", sd(stationary$oir, na.rm = TRUE), "(should be ~1)\n")

# Create ML dataset with proper train/test split
ml_features <- data.frame(
  feature1 = rnorm(100),
  feature2 = rnorm(100),
  feature3 = rnorm(100),
  target = rnorm(100)
)

dataset <- create_ml_dataset(
  ml_features,
  target_col = "target",
  train_fraction = 0.70
)

cat("Train set:", nrow(dataset$train), "rows\n")
cat("Validation set:", nrow(dataset$validation), "rows\n")
cat("Test set:", nrow(dataset$test), "rows\n")
```

## Complete Workflow Example

Here's a complete analysis using all modules:

```{r workflow}
# 1. Generate test data
test_data <- generate_test_data(n_trades = 500, seed = 123)

# 2. Basic OFI
ofi <- compute_ofi(test_data$trades, window = "1 min")

# 3. Visualize
plot_ofi_diagnostics(ofi)

# 4. Test for manipulation
alerts <- surveillance_alert_system(test_data$messages)
cat("Alerts generated:", nrow(alerts$alerts), "\n")

# 5. Multi-level metrics
ml_metrics <- compute_multilevel_ofi(test_data$orderbook, n_levels = 5)
cat("Multi-level OFI improvement:", ml_metrics$improvement_vs_single_level, "%\n")

# 6. Price impact
impact <- sqrt_impact(Q = 50000, V = 1000000, sigma = 0.02)
cat("Expected impact:", round(impact, 2), "bps\n")

# 7. Feature engineering for ML
features <- engineer_ofi_features(test_data$trades, lookback = 5)
cat("ML features generated:", ncol(features), "\n")
```

## Key Takeaways

1. **generate_test_data()** creates all data formats you need
2. Each module has specific data requirements (trades vs messages vs orderbook)
3. You can test **every** function without real market data
4. Synthetic data is perfect for learning and development
5. When you're ready, apply the same functions to real data!

## Next Steps

- Try each module with your own parameters
- Experiment with different data generation settings
- When ready, load real data using `read_taq_trades()` or `read_lobster_trades()`
- The same functions work on both synthetic and real data!

## Additional Resources

```{r, eval=FALSE}
# Explore all functions
?generate_test_data
?compute_ofi
?detect_spoofing
?compute_multilevel_ofi

# See all available vignettes
browseVignettes("rOFI")
```
